Supplemental Materials: Math Glossary

### 2025-11-27 paper/main.tex
accessviz submission
fold back in equivariance and homomorphism

\usepackage{silence}
\WarningsOff[fixltx2e]

@@

% *** MATH PACKAGES ***
%% with other math-related packages, you may want to disable it.
\usepackage{amsmath, amsthm, amsfonts,amssymb,eulervm,xspace, mathtools}
\usepackage{stmaryrd}%maps from
%\renewcommand{\restriction}{\mathord{\upharpoonright}} %restriction w/p space
\usepackage{mathrsfs} % math script fonts
\usepackage{relsize} %bigger
\usepackage{bm}

@@

\usepackage[]{minted}
\setminted[python]{fontsize=\scriptsize,
                   linenos,
                   numbersep=8pt,
                   frame=lines,
                   autogobble,
                   framesep=3mm,
                   breaklines=True}
% *** ALIGNMENT PACKAGES ***
\usepackage{multicol}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabulary}
\usepackage{tabularray}
% IEEEtran contains the IEEEeqnarray family of commands

@@

We propose that we can better enforce this expectation in software by providing a uniform way of expressing $data$ and $graphic$ using their respective algebraic structure and by uniformly specifying the behaviors and properties of those structures and the maps between them using category theory. For example, our framework can encapsulate how a table and scatter plot and heatmap are different representations of the same data and track an observation from a data cube as a point along a time series and on a map and in a network. The algebraic structures can then be translated into programmatic types, while the categorical descriptions translate to a functional design framework. Strong typing and function composition enable visualization software developers to build complex components from simpler verifiable parts \cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These components can be built as a standalone library and integrated into existing libraries and we hope these ideas will influence the architecture of critical data visualization libraries, such as Matplotlib.

@@

\begin{LaTeXdescription}
  \item[\textcolor{fiber}{\textbf{field}}]\footnote{In this paper, color in definitions, equations, and visualizations is used to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}.} is a set of values of the same type, e.g. one column of a table or the pixels of an image
  \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}.
\end{LaTeXdescription}

@@

Data is often described by its mathematical structure, for example the Steven's measurement scales define nominal, ordinal, interval, and ratio data by the allowed operations on each \cite{stevensTheoryScalesMeasurement1946} and other researchers have since expanded the scales to encapsulate more types of structure \cite{leaFormalizationMeasurementScale1971, thomasMathematizationNotMeasurement2014}. Loosely we define structure as a mathematical space $X$ and the actions on that space; our model is purposely generalizable to many different types of spaces (e.g sets, metric spaces, etc). Steven's conceptualizes the scales as mathematical groups with \textcolor{action}{actions}. A function that preserves structure when the input or output is changed by a group action is called \textit{equivariant}; a visualization is structure preserving when there exist compatible group actions on the data and visualization was proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}. Although the Steven's scales were conceptualized as having group structure, the ordinal scale has a monoidal structure because partial orders ($\geq, \leq$) are not invertible. This means \textit{equivariance} cannot be used to test for structure preservation; instead \textit{homomorphism} can be used to verify whether map preserve relations between data elements, as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}. These tests are mutually exclusive; a function can be homomorphic but not equivariant, such as an exponential encoding, or equivariant but not homomorphic, such as the inverse encoding. A function can also be homomorphic (or equivariant) with respect to one action but not with respect to another. The encoding transforms in visualization tools are expected to preserve the structure of whatever input they receive; therefore a methodology for codifying arbitrary structure is presented in \autoref{sec:atct:fb:fiber} and \autoref{sec:artist:equivariant:data} presents a generalization of equivariance and homomorphism for evaluating structure preservation.

@@

\begin{LaTeXdescription}
  \item[pan, zoom, scroll] sheaf: locality + gluing \autoref{def:atct:sheaf}
  \item[selection and hover] pushforward \autoref{eq:atct:sheaf:pushforward_select},  pullback \autoref{eq:atct:sheaf:pullback_hover}
  \item[brushing, linking, annotation] composition of artists \autoref{eq:artist:join_base_fiber}
\end{LaTeXdescription}

@@

Refer to the instructions for this section (\cref{sec:supplement_inst}).
Below is an example you can follow that includes the actual supplemental material for this template:

All supplemental materials are available on OSF at \url{https://doi.org/10.17605/OSF.IO/2NBSG}, released under a CC BY 4.0 license.
In particular, they include (1) Excel files containing the data for and analyses for creating \cref{tab:vis_papers} and \cref{fig:vis_papers}, (2) figure images in multiple formats, and (3) a full version of this paper with all appendices.
Our other code is intellectual property of a corporation---Starbucks Research---and there is no feasible way to share it publicly.

\appendices

\section{Summary}
\label{sec:appndix:summary}
The topological spaces and functions introduced throughout this paper are summarized here for reference.

@@

\section*{Acknowledgment}
\note{Acknowledge all the actual people}
The authors would like to thank the anonymous reviewers who gave constructive feedback on an earlier version of this paper. The authors are also grateful to  the various Matplotlib and Napari contributors, particularly Juan Nunez-Iglesias, and Nicolas Kruchten for their valuable feedback from the library developer perspective.

This project has been made possible in part by grant number 2019-207333 and \note{Cycle 3 grant number} Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
  \newpage
\fi

% references section
\bibliographystyle{abbrv-doi}
\bibliography{references}
\end{document}

>>>>

%\usepackage{silence}
%\WarningsOff[fixltx2e]

@@

% *** MATH PACKAGES ***
%% with other math-related packages, you may want to disable it.
\usepackage{mathptmx}
\usepackage{amsmath, amsthm, amsfonts,amssymb,eulervm,xspace, mathtools}
\usepackage{stmaryrd}%maps from
%\renewcommand{\restriction}{\mathord{\upharpoonright}} %restriction w/p space
\usepackage{mathrsfs} % math script fonts
\usepackage{relsize} %bigger
\usepackage{bm}

@@

\usepackage[cache=false]{minted}
\setminted[python]{fontsize=\scriptsize,
                   linenos,
                   numbersep=8pt,
                   frame=lines,
                   autogobble,
                   framesep=3mm,
                   breaklines=True}
% *** ALIGNMENT PACKAGES ***
\usepackage{multicol}
\usepackage{array}
\usepackage{multirow}
\usepackage{tabulary}
\usepackage{tabularray}
% IEEEtran contains the IEEEeqnarray family of commands

@@

We propose that structures from category theory and algebraic topology can be used as a sort of psuedo-code for uniformly specifying the behaviors and properties of $data$ and $graphic$ and rigorously expressing the structure preserving constraints on maps between them. For example, our framework can encapsulate how a table and scatter plot and heatmap are different representations of the same data and track an observation from a data cube as a point along a time series and on a map and in a network. By design, this psuedo-code translates to strong programmatic types and a functional design framework, which enable visualization software developers to build complex components from simpler verifiable parts \cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These components can be built as a standalone library and integrated into existing libraries and we hope these ideas will influence the architecture of critical data visualization libraries, such as Matplotlib.

@@

\begin{description}
  \item[\textcolor{fiber}{\textbf{field}}]\footnote{In this paper, color in definitions, equations, and visualizations is used to group conceptually related terms\cite{headMathAugmentationHow2022}, for example \textcolor{fiber}{field} and \textcolor{fiber}{fiber}.} is a set of values of the same type, e.g. one column of a table or the pixels of an image
  \item[\textcolor{base}{\textbf{topology}}] is the connectivity and relative positioning of elements in a dataset \cite{wilkinsonGrammarGraphics2005}.
\end{description}

@@

Data is often described by its mathematical structure, for example the Steven's measurement scales define nominal, ordinal, interval, and ratio data by the transformations allowed on the data \cite{stevensTheoryScalesMeasurement1946}. Other researchers have since expanded the scales to encapsulate more types of structure \cite{leaFormalizationMeasurementScale1971, thomasMathematizationNotMeasurement2014}, and in \autoref{sec:atct:fb:fiber} we generalize further by proposing that data can be encoded as mathematical categories.

Loosely, structure is defined as a collection of values $X$ and transformations $G$ on those values. These transformations can be operations, relations, or generalized as actions:
\begin{definition}\label{def:related-work:action}
  An \textcolor{action}{\textbf{action}} is a function  $act: \textcolor{action}{G} \times X \rightarrow X$. An action has the properties of identity $act(\textcolor{action}{e}, x) = x$ for all  $x \in X$ and associativity $act(\textcolor{action}{g}, act(\textcolor{action}{f}, x)) = act(\textcolor{action}{f} \circ \textcolor{action}{g}, x)$ for $\textcolor{action}{f},\textcolor{action}{g} \in \textcolor{action}{G}$.\cite{grimaldiDiscreteCombinatorialMathematics2006}
\end{definition}
Elements of $X$ can be from one data field or all of them or some subset; similarly the actions act on the elements of $X$ and each action can be a composition of actions.

A function that preserves structure when the input or output is changed by a group\footnote{A mathematical {group} is a set with an associative binary operator. This operation must have an identity element and be closed, associative, and invertible, consisting of a set of values $X$ and \textcolor{action}{actions} on the set {$G = (G,\circ, e)$}.} action is called \textit{equivariant}.  Given a group $G$ that acts on both $X$ and $Y$, 
\begin{definition}\label{def:equivariance}
 a function $f: X \rightarrow Y$ is \textbf{equivariant} when $f(act(g,x)) = act(g,f(x))$ for all $g$ in $G$ and for all $x$ in $X$ \cite{pittsNominalSetsNames2013}
\end{definition}
This means that a visualization is structure preserving when there exist compatible group actions on the data and visualization, as proposed by Kindlemann and Scheidegger\cite{kindlmannAlgebraicProcessVisualization2014}.

Although the Steven's scales were conceptualized as having group structure, the ordinal scale has a monoidal structure because partial orders ($\geq, \leq$) are not invertible. This means \textit{equivariance} cannot be used to test for structure preservation; instead \textit{homomorphism} can be used. Given the function $f: X \rightarrow Y$, with operators $(X, \circ)$ and $(Y, *)$
\begin{definition}\label{def:homomorphism}
  A function $f$ is \textbf{homomorphic} when $f(x_1 \circ x_2) = f(x_1) * f(x_2)$ and preserves identities $f(I_x) = I_y$ for all $x, y \in X$ \cite{grimaldiDiscreteCombinatorialMathematics2006}
\end{definition}
This means that a visualization is structure preserving when there
exists compatible operators (here $\circ$ and $*$) such that the relations between data elements is preserved in the mapping, 
as proposed by Mackinlay\cite{mackinlayAutomaticDesignGraphical1987}.

\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{equiv_homo.pdf}
  \caption{Encoding data as the bar height using an exponential transform is not equivariant because encoding the data and then scaling the bar heights yields a much taller graph then scaling the data and then encoding those heights using the same exponential transform function. Encoding data as bar height  using an inverse transform is not homomorphic because the largest number is mapped to the smallest bar while the max function returns the largest bar.}
  \label{fig:related-work:field-preservation}
\end{figure}

As illustrated in \autoref{fig:related-work:field-preservation}, a visualization is structure preserving when the functions commute such that the graphic is the same whether the data is acted on ($
\dfunct$) and then mapped ($A$) to a graphic or whether the data is mapped to a graphic that is then modified in a compatible way. As shown in \autoref{fig:related-work:field-preservation}, a function can be homomorphic but not equivariant, such as an exponential encoding, or equivariant but not homomorphic, such as the inverse encoding. As shown, a function can also be homomorphic (or equivariant) with respect to one action but not with respect to another. We generalize the expectation of commutativity to codify structure preservation in \autoref{sec:artist:equivariant:data}.

@@

\begin{description}
  \item[pan, zoom, scroll] sheaf: locality + gluing \autoref{def:atct:sheaf}
  \item[selection and hover] pushforward \autoref{eq:atct:sheaf:pushforward_select},  pullback \autoref{eq:atct:sheaf:pullback_hover}
  \item[brushing, linking, annotation] composition of artists \autoref{eq:artist:join_base_fiber}
\end{description}

@@

\acknowledgments{
The authors would like to thank the anonymous reviewers who gave constructive feedback on an earlier version of this paper. The authors are also grateful to  the various Matplotlib and Napari contributors, particularly Juan Nunez-Iglesias, and Nicolas Kruchten for their valuable feedback from the library developer perspective.

This project has been made possible in part by grant number 2019-207333 and \note{Cycle 3 grant number} Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation
}

% references section
\bibliographystyle{abbrv-doi}
\bibliography{references}

\appendix

\section{Summary}
\label{sec:appendix:summary}
The topological spaces and functions introduced throughout this paper are summarized here for reference.

@@

\end{document}

### 2025-11-27 paper/supplement.tex
fold back in equivariance and homomorphism

% Fonts
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{stmaryrd}%maps from
\renewcommand{\restriction}{\mathord{\upharpoonright}} %restriction w/p space
\usepackage{mathrsfs} % math script fonts
\usepackage{mathptmx}
\DeclareMathDelimiter{(}{\mathopen} {operators}{"28}{largesymbols}{"00}
\DeclareMathDelimiter{)}{\mathclose}{operators}{"29}{largesymbols}{"01}
% *** SUBFIGURE PACKAGES ***
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}

@@

\subsection{Natural transforms}
\begin{definition}\label{def:natural-transform}
  Given two functors $F, G: \mathcal{C}\rightarrow \mathcal{D}$, a \textbf{natural transformation} $\alpha: F \rightarrow G$ is a function which assigns to each object $c$ of $\mathcal{C}$ a morphism $\alpha_c:F(input) \rightarrow G(c), G(c) \in \mathcal{D}$, in such a way that for every morphism $f:c \rightarrow c^\prime, c^\prime \in \mathcal{C}$, the morphisms in $\mathcal{D}$ commute such that $\alpha_c^{\prime}(F(f)(F(c))) = G(f)(\alpha_c(F(c))$. When this holds, $\alpha_{c}$ is \textit{natural} in $c$.\cite{maclaneCategoriesWorkingMathematician2013}.
\end{definition}.

@@

\section{Presheaves and Sheaves}
A common way of encapsulating a map from an arbitrary category to a category of sets is as a presheaf and sheaves add on conditions for gluing individual sections over subspaces into cohesive sections over the whole space.
\subsection{Presheaf}
\begin{definition}\label{def:atct:presheaf}
  A \textbf{presheaf} $F:\mathcal{C}^{op} \rightarrow \setb$ is a contravariant functor from an object in an arbitrary category to an object in the category \setb\cite{spanier1989algebraic}.
\end{definition}

@@

\subsection{Germ}
While each section of a sheaf is evaluated over a point $\dsection(\dbasepoint)$ such that it returns a single record, the sheaf model also provides an abstraction when neighboring information is required. The sheaf over a very small region surrounding a point $\dbasepoint$ is called a \textit{stalk}\cite{harder2008lectures}
\begin{equation}
  \label{eq:atct:sheaf:stalk}
    \sheaf_{\dbase, \dtotalc}\restriction_{\dbasepoint}\coloneqq \lim\limits_{\openset\ni \dbasepoint} \Gamma(\openset, \dtotal\restriction_{\openset})
\end{equation}
where the fiber is contained inside the stalk  $\dfiber_{\dbasepoint} \subset  \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. The \textit{germ} is the section evaluated at a point in the stalk  $\dsection(\dbasepoint) \in \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. Since the stalk includes the values near the limit of the point at \dbasepoint\, the germ can be used to compute mathematical derivatives.

>>>>

% Fonts
% Fonts: use T1 encoding and a Times-like text+math package for pdfLaTeX
\usepackage[T1]{fontenc}
\usepackage{newtxtext,newtxmath}
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{stmaryrd}%maps from
\renewcommand{\restriction}{\mathord{\upharpoonright}} %restriction w/p space
\usepackage{mathrsfs} % math script fonts
% *** SUBFIGURE PACKAGES ***
\usepackage[caption=false,font=normalsize,labelfont=sf,textfont=sf]{subfig}

@@

\subsection{Natural transforms}
\begin{equation*} {\renewcommand{\arraystretch}{1.2}
  \begin{tabular}{|c|c|}
    \hline
    $\mathcal{C}$ & $\mathcal{D}$ \\
    \hline
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
        \textcolor{base}{c} \arrow[d, "f", color=action] \\ \textcolor{base}{c^{\prime}}      
    \end{tikzcd}
    & 
    \begin{tikzcd}[ampersand replacement=\&, sep=huge]
        \textcolor{sheaf}{F(c)} \arrow[d, "F(f)", color=action] \arrow[r, "\alpha_{c}", color=artist] \& 
        \textcolor{sheaf}{G(c)} \arrow[d, "G(f)", color=action] \\
        \textcolor{sheaf}{F(c^{\prime})} \arrow[r, "\alpha_{c^{\prime}}", color=artist] \& \textcolor{sheaf}{G(c^{\prime})} 
    \end{tikzcd}\\
    \hline
  \end{tabular}}
\end{equation*}

\begin{definition}\label{def:natural-transform}
  Given two functors $F, G: \mathcal{C}\rightarrow \mathcal{D}$, a \textbf{natural transformation} $\alpha: F \rightarrow G$ is a function which assigns to each object $c$ of $\mathcal{C}$ a morphism $\alpha_c:F(c) \rightarrow G(c)$ of $\mathcal{D}$, in such a way that for every morphism $f:c \rightarrow c^\prime$ in $\mathcal{C}$, the morphisms $F(f), G(f)$ in $\mathcal{D}$ commute such that $\alpha_c^{\prime}(F(f)(F(c))) = G(f)(\alpha_c(F(c)))$. When this holds, $\alpha_{c}$ can be said to be \textit{natural} in $\mathcal{C}$.\cite{maclaneCategoriesWorkingMathematician2013}.
\end{definition}.

@@

\section{Presheaves and Sheaves}
A common way of encapsulating a map from an arbitrary category to a category of sets is as a presheaf and sheaves add on conditions for gluing individual sections over subspaces into cohesive sections over the whole space.
\subsection{Presheaf}

\begin{definition}\label{def:atct:presheaf}
  A \textbf{presheaf} $F:\mathcal{C}^{op} \rightarrow \setb$ is a contravariant functor from an object in an arbitrary category to an object in the category \setb\cite{spanier1989algebraic}.
\end{definition}

\begin{equation*}
\begin{tikzcd}
\cgamma{\opensetc_2}{\dtotalc|_{\opensetc_2}} \arrow[r, "
\hat{\phi}^*", color=action, dashed] & \cgamma{\opensetc_1}{\dtotalc|_{\opensetc_1}}                \\
\opensetc_2 \arrow[u, "\sheaf_{2}", maps to, color=sheaf]                  & \opensetc_1 \arrow[l, "\hat{
  \phi}"', color=action] \arrow[u, "\sheaf_{1}", maps to, color=sheaf]
\end{tikzcd}
\end{equation*}

@@

\subsection{Germ}
While each section of a sheaf is evaluated over a point $\dsection(\dbasepoint)$ such that it returns a single record, the sheaf model also provides an abstraction when neighboring information is required. The sheaf over a very small region surrounding a point $\dbasepoint$ is called a \textit{stalk}\cite{harder2008lectures}
\begin{equation}
  \label{eq:atct:sheaf:stalk}
    \sheafc_{\dbasec, \dtotalc}\restriction_{\dbasepointc}\coloneqq \lim\limits_{\opensetc\ni \dbasepointc} \Gamma(\opensetc, \dtotalc\restriction_{\opensetc})
\end{equation}
where the fiber is contained inside the stalk  $\dfiber_{\dbasepoint} \subset  \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. The \textit{germ} is the section evaluated at a point in the stalk  $\dsection(\dbasepoint) \in \sheaf_{\dbase, \dtotal}\restriction_{\dbasepoint}$. Since the stalk includes the values near the limit of the point at \dbasepoint\, the germ can be used to compute mathematical derivatives.
